Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-08 Manuscript/MPESubmission/Revision/response.tex

We modified the following sentence as suggested: "as long as enough data overlaps in each case, and missing data are not phylogenetically biased" (line @@@).

\item{\textcolor{blue}{\textbf{line 99:} unclear - what is meant by 'with a fixed topology approach'? The studies references investigated divergence times using total evidence datasets and in some cases used constraints but not completely fixed topologies}}

We modified the previous statement to: "but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints" (line @@@).

\item{\textcolor{blue}{\textbf{lines 80-103:} (1) the authors need to clearly justify the importance of this study compared the many studies that have investigated the effects of missing data on topology inference.
The justification currently given is not particularly strong and the specific unique questions asked in this study that are currently unanswered should be made clear.
The current argument does not always seem justified by the literature review.
For example, line 81 the authors argue that papers by Wiens and Pattinson examined morphological or molecular matrices separately, but in Pattinson, for example, combined datasets for living species were used to simulate extinct species by deleting the molecular and parts of the morphological datasets from living species.
The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
That is one aspect of this study that I perceived as new and unique.
(2) Another aspect of this study design that could be new but is not explored here is the effect of the missing data on the morphological substitution rate parameter estimates.
Can you recover the true values from the perturbed datasets?
This study also used a unique way of evaluating the tree inference under missing data scenarios.
These and other aspects of the study uniqueness must be clearer.}}

(1) - We thank the reviewer for this judicious comment and emphasized on both the novelty aspect of our study and on the questions we are actually testing as follow:

"Until now, only few attempts have been made to study the impact of this issue on phylogenetic inference using Total Evidence methods (e.g. Wiens et al., 2005; Pattinson et al., 2014).%\citep[e.g.][]{Wiens01102005,pattinsonphylogeny2014}
For example, Wiens et al. (2005) %\cite{Wiens01102005}
 only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.
Furthermore, Pattinson et al. (2014) %\cite{pattinsonphylogeny2014}
 assess the effect of removing both molecular and morphological data (``artificial extinction'') on the topology.
These studies, however, are limited by their empirical approach (Wiens et al., 2005) and the scheme of generating missing data (Pattinson et al., 2014).
In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by thoroughly eliminating living or fossil data or all data overall.
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa."
%Pattinson: artificial fossil - removing molecular data from living species to see how topology changes. Pattinson use combined analysis but only to understand the patterns of fossilisation (via artificial fossil).
%Wiens only focuses on a special case
%Limitation of Pattinson was the limited scheme for generating missing data from the datasets (specific to primates and observed patterns).
%In our case, we (a) don't focus on any group or any pattern, our approach is purely theoretical.
%The difference from Wiens or Pattinson is that we thoroughly investigate the effects of eliminating living versus fossil data or all data overall.
%(b) We also propose a thorough comparison between the ML and Bayesian inference and a measure two clear effects that missing data could have on topology: displacing wild-card taxa and not conserving clades structure.

@@

%Emphasize on the questions and the uniqueness
%The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
%In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
%That is one aspect of this study that I perceived as new and unique.
%This study also used a unique way of evaluating the tree inference under missing data scenarios.

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting. However, we considered this study to be already long and complex. We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might loose the reader.
Nonetheless, we added the result of such analysis in the \textbf{Appendix-A Morphological rates estimations results} section.
% TG: what do you think? I don't want to attract the reviewer's attention too much (i.e. that's another shit load of analysis to do to do it properly) and certainly not the reader (it's not relevant to the whole paper) but sticking that to the supplementary kind of does what the reviewers want to know.

### 2015-07-08 Manuscript/MPESubmission/Revision/OnlineAppendices-LaTeXSuppFiles/AppendixA.tex
Revision init
Added rates analysis to appendix A

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}

@@

@@

\begin{itemize}
  \item Molecular data: GTR + $\Gamma_4$ (-m GTRGAMMA)
  \item Morphological data: Mk + $\Gamma_4$ (-K MK)
  \item Support: Rapid Boostrap algorithm (LSR), 1000 replicates
\end{itemize}

@@

\begin{itemize}
  \item Priors: Molecular data
  \begin{itemize}
    \item Rates distribution shape ($\alpha$) = 0.5
    \item Transition/Transversion ratio = 2 ($\beta$(80,40))
    \item Starting tree: "True" tree topology with each branch length = 1
  \end{itemize}
  \item Priors: Morphological data
  \begin{itemize}
    \item rates distribution shape ($\alpha$) = 0.5
  \end{itemize}
  \item Models
  \begin{itemize}
    \item Molecular data: HKY + $\Gamma_4$
    \item Morphological data: Mk + $\Gamma_4$
  \end{itemize}
  \item MCMC
  \begin{itemize}
    \item Two runs
    \item Four chains per run
    \item Generations $<$ 5$\times$$10^7$
    \item Sample frequency = 1.05$\times$$10^4$
    \item ASDS diagnosis frequency = 5$\times$$1^4$
    \item ASDS $<$ 0.01
    \item ESS $>>$ 200
    \item Burnin = 25\%
  \end{itemize}
\end{itemize}

>>>>

%Packages
\usepackage{xcolor}
\usepackage{changepage}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{longtable}
\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
%\setcounter{secnumdepth}{0}

@@

\subsection{Morphological rates estimations results}
To test the sensitivity of the Bayesian prior on the morphological rates distribution shape ($\alpha$ = 0.5), we compared the shape of the distribution of our generated morphological characters ($\alpha_{TRUE}$) to the shape of the distribution of morphological characters rate estimated by the Bayesian inference ($\alpha_{EST}$).
The $\alpha_{TRUE}$ parameter was estimated by fitting a gamma distribution to the 100 individual morphological rates (randomly sampled from a gamma distribution with $\alpha$ = 0.5) using the \texttt{fitdistr} function from the \texttt{MASS R} package \citep{MASS}.
We measured the difference as:
\begin{equation}
\Delta\alpha=\alpha_{TRUE} - \alpha_{EST}
\end{equation}
We then calculated the mode and the 95\% and 50\% confidence intervals over the 50 chains using the \texttt{hdr} function from the \texttt{hdrcde R} package \citep{hdrcde}.
The code used for this section is available at \url{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data/blob/master/Analysis/Rates_estimates.R}.

\begin{figure}
\centering
\includegraphics[width=\textwidth,keepaspectratio]{SupplementaryFigures/Rates_estimates.pdf}
\caption{Differences between the simulated rate of each morphological character ($\alpha$ parameter) and the Bayesian estimates for the trees with missing data. The x axis shows the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$ (lower line). Positive values shows underestimation of the parameter and negative values shows overestimation. One unit of difference means that the estimated rate is twice as big/small than the true rate. The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the rates estimates.}
\label{Fig_AppendixCharacters}
\end{figure}

@@

\begin{itemize}
  \item Molecular data: GTR + $\Gamma_4$ (-m GTRGAMMA)
  \item Morphological data: M\textit{kv} + $\Gamma_4$ (-K MK)
  \item Support: Rapid Boostrap algorithm (LSR), 1000 replicates
\end{itemize}

@@

\begin{itemize}
  \item Priors: Molecular data
  \begin{itemize}
    \item Rates distribution shape ($\alpha$) = 0.5
    \item Transition/Transversion ratio = 2 ($\beta$(80,40))
    \item Starting tree: "True" tree topology with each branch length = 1
  \end{itemize}
  \item Priors: Morphological data
  \begin{itemize}
    \item rates distribution shape ($\alpha$) = 0.5
  \end{itemize}
  \item Models
  \begin{itemize}
    \item Molecular data: HKY + $\Gamma_4$
    \item Morphological data: M\textit{kv} + $\Gamma_4$
  \end{itemize}
  \item MCMC
  \begin{itemize}
    \item Two runs
    \item Four chains per run
    \item Generations $<$ 5$\times$$10^7$
    \item Sample frequency = 1.05$\times$$10^4$
    \item ASDS diagnosis frequency = 5$\times$$1^4$
    \item ASDS $<$ 0.01
    \item ESS $>>$ 200
    \item Burnin = 25\%
  \end{itemize}
\end{itemize}
